---
title: "Quiz 3 Solutions"
author: "Daniel Iong"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(reticulate)
```

Use Monte-Carlo sampling to answer the following question:

Let $X_1, \dots, X_n$ be Exponential(5) random variables and $\bar{X}_n = \frac{1}{n} \sum_{i=1}^n X_i$. If $n$ is large,

1. What is $E[\log(\bar{X}_n)]$ (approximately)? 
2. What is the (approximate) distribution of $\sqrt{n} \big[ \log(\bar{X}_n) - E[\log(\bar{X}_n)] \big]$? 


```{r, cache=TRUE}
exp.rate <- 5 # Exponential distribution rate
N.mc <- 50000 # Number of Monte-Carlo samples 
N.samples <- 100000
xbar.vec <- c() # Vector containing sample means 
for (i in 1:N.mc) {
  exp.samples <- rexp(N.samples, exp.rate) # Sample from Exp(exp.rate) N.samples times
  samp.mean <- mean(exp.samples) # Calculate sample mean 
  xbar.vec <- c(xbar.vec, samp.mean) # append sample mean to xbar.vec
}
```

```{r}
# 1.
log_xbar.mean <- mean(log(xbar.vec)) # Approximate distribution mean
log_xbar.mean

# 2.
hist(sqrt(N.samples) * (log(xbar.vec) - log_xbar.mean)) # Histogram of log of sample means 
var(sqrt(N.samples) * (log(xbar.vec) - log_xbar.mean)) # Approximate distribution variance
```


**Remarks**

* This question should give you an idea of the level of R programming required for this course. The question is conceptually difficult but not too difficult to code.

----

**NOT REQUIRED FOR THIS COURSE**: This question can also be answered analytically using the [delta method](https://en.wikipedia.org/wiki/Delta_method). 

$E[\log(\bar{X}_n)] = \log(1/5) =$ `r log((1/5))`

$$\sqrt{n} \big[ \log(\bar{X}_n) - E[\log(\bar{X}_n)] \big] \overset{\mathcal{D}}{\rightarrow} N \Big(0, \sigma^2 [\log(\mu)']^2 \Big) = N \Big( 0, \frac{1}{5^2} * 5^2 = 1 \Big)$$





# Python Solution





