---
title: "Quiz 3 Solutions"
author: "Daniel Iong"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Quiz 3

Use Monte-Carlo sampling to answer the following question:

Let $X_1, \dots, X_n$ be Exponential(5) random variables and $\bar{X}_n = \frac{1}{n} \sum_{i=1}^n X_i$. If $n$ is large,

1. What is $E[\log(\bar{X}_n)]$ (approximately)? 
2. What is the (approximate) distribution of $\sqrt{n} \big[ \log(\bar{X}_n) - E[\log(\bar{X}_n)] \big]$? 


```{r, cache=TRUE}
exp.rate <- 5 # Exponential distribution rate
N.mc <- 50000 # Number of Monte-Carlo samples 
N.samples <- 100000
xbar.vec <- c() # Vector containing sample means 
for (i in 1:N.mc) {
  exp.samples <- rexp(N.samples, exp.rate) # Sample from Exp(exp.rate) N.samples times
  samp.mean <- mean(exp.samples) # Calculate sample mean 
  xbar.vec <- c(xbar.vec, samp.mean) # append sample mean to xbar.vec
}
```

```{r, echo = FALSE}
xbar.mean_exact <- log((1/5))
```

```{r}
# 1.
log_xbar.mean <- mean(log(xbar.vec)) # Approximate distribution mean
log_xbar.mean

# 2.
hist(log(xbar.vec)) # Histogram of log of sample means 
var(sqrt(N.samples) * (log(xbar.vec) - log_xbar.mean)) # Approximate distribution variance
```

This question can also be answered analytically using the delta method. 

$E[\log(\bar{X}_n)] = \log(1/5) =$ `r log((1/5))`

$$\sqrt{n} \big[ \log(\bar{X}_n) - E[\log(\bar{X}_n)] \big] \Rightarrow N \Big(0, \sigma^2 [\log(\mu)']^2 \Big) = N \Big( 0, \frac{1}{5^2} * 5^2 = 1 \Big)$$
**Answer Choices** 


a. $E[\log(\bar{X}_n)] = -1.61$, $\sqrt{n} \big[ \log(\bar{X}_n) - E[\log(\bar{X}_n)] \big] \sim N(0, 1)$

b. $E[\log(\bar{X}_n)] = 0.62$, $\sqrt{n} \big[ \log(\bar{X}_n) - E[\log(\bar{X}_n)] \big] \sim N(0, 1)$

c. $E[\log(\bar{X}_n)] = -1.61$, $\sqrt{n} \big[ \log(\bar{X}_n) - E[\log(\bar{X}_n)] \big] \sim N(0, 0.04)$

d. $E[\log(\bar{X}_n)] = 0.62$, $\sqrt{n} \big[ \log(\bar{X}_n) - E[\log(\bar{X}_n)] \big] \sim N(0, 0.04)$

